Vertical and oblique cultural transmission fluctuating in time and in space

### 2018-11-07 ms/ms_sunnyvale.tex
submitted revision to tpb

% line numbers
%\usepackage[displaymath, mathlines]{lineno}
%\renewcommand\linenumberfont{\normalfont\small\sffamily}
%\linenumbers
%\modulolinenumbers[2]

@@

\author[a]{Yoav Ram}
\author[b]{Uri Liberman}
\author[a]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{School of Mathematical Sciences, Tel Aviv University, Israel}

@@

Suppose that both transmission and selection fluctuate together, so that when $A$ is favored, with $w_A=1+s$ and $w_B=1$ with $s>0$, the transmission rate is $\rho_A$, and when $B$ is favored, with $w_A=1$ and $w_B=1+s$, the transmission rate is $\rho_B$.
The change in the frequency $x$ of phenotype $A$ when either $A$ or $B$ is favored is described by $F_A(x)$ or $F_B(x)$, respectively, where
\begin{equation}\begin{aligned} \label{eq:recurrence_periodic_fluc}
&F_A(x) = \rho_A \frac{1+s}{1+sx} x + (1-\rho_A)x, &\quad \text{and} \\
&F_B(x) = \rho_B \frac{1}{1+s-sx} x + (1-\rho_B)x.
\end{aligned}\end{equation}

@@

% Period of one
\paragraph{Period of one---$A1B1$.}
In the case  $k=1$ we can find $x^*$ the stable frequency of $A$ at the protected polymorphism.
Set $F_B(F_A(x))-x = G(x) \cdot x (1-x) \cdot \gamma$ with
\begin{equation} \begin{aligned} \label{eq:xstar_periodic_fluc_k=1}
&G(x) = ax^2+bx+c, \\
&\gamma = -\frac{1 + sx}{s} \Big[ 1 + s + s^2 x (1-\rho)(1+x)\Big], \\
&a = s^2 \rho_A (1-\rho_A) (1 - \rho_B + \rho_B/\rho_A)) > 0, \\
&b = s(1-\rho_A)(2\rho_B - s\rho_A(1-\rho_B)), \\
& c = \rho_B - \rho_A - s\rho_A(1-\rho_B).
\end{aligned} \end{equation}
Then $x^*$ is a solution of the quadratic $G(x)=0$.
The condition \eqref{eq:poly_condition_periodic_fluc}, which guarantees that $0 < x^* < 1$, is equivalent to $c<0$, and therefore it also guarantees that $\sqrt{b^2-4ac} > b$.
So, if $0<s<1$ (so that $b$ is guaranteed to be positive if LHS of eq.~\ref{eq:poly_condition_periodic_fluc} holds), then 
\begin{equation}
x^*= \frac{-b+\sqrt{b^2-4ac}}{2a}.
\end{equation}
\autoref{fig:rho1_rho2_phase_k=1} shows $x^*$ and highlights the area of the parameter space in which a protected polymorphism exists.
The figure demonstrates that the stronger the selection (x-axis), the greater the fluctuations in $\rho$  (y-axis) that still allow a polymorphic population (area between the dashed lines).

@@

% Asymmetric period
\paragraph{Asymmetric period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
& 1 < (F'_A(0))^k (F'_B(0))^l = a^k \big(b - s \Delta \rho /(1+s)\big)^l, \quad \text{and} \\
& 1 < (F'_A(1))^k (F'_B(1))^l = b^k (a + s \Delta \rho)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits a protected polymorphism; decreasing $\rho_B$  will have the opposite effect: it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Second, if $s_t$ and $\rho_t$ are not independent such that $cov(\rho_t, s_t) \ne 0$, a fixation can occur.
Let $w_{A,t}, w_{B,t} \sim U(0,1)$, $s_t=(w_{A,t}-w_{B,t})/w_{B,t}$ and $\rho_t \sim Beta(1+s_t, 1)$ (a beta distribution with parameters $1+s_t$ and $1$). 
The covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 4$ as estimated by averaging over $10^8$ random values of $s_t$ and $\rho_t$); that is, vertical transmission is more likely when $A$ is favored (i.e. $s_t>0$) and oblique transmission is more  likely when $B$ is favored (i.e. $s_t<0$; \autoref{fig:beta}).
Then $E[log(1+z_t)] >0$ and $B$ almost surely doesn't fix. 
Also, $E[log(1-z_t / (1+s_t)] < 0$, so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1, 1+s_t)$ and the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -4$). In that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=\linewidth]{../figures/{beta}.pdf}
\caption{
\textbf{Covariance of selection and transmission.}
\textbf{(a)} Histogram of $w_{A,t}/w_{B,t}$ where $w_{A,t}$ and $w_{B,t}$ are identically and independently distributed uniform random variables $U(0,1)$.
\textbf{(b)} Histogram of $s_t = (w_{A,t}-w_{B,t})/w_{B,t}$.
\textbf{(c)} Histogram of $\rho_t \sim Beta(1+s_t, 1)$.
\textbf{(d)} The joint distribution of $\rho_t$ and $s_t$ demonstrates a positive correlation $cov(s_t, \rho_t)>0$.
}
\label{fig:beta}
\end{figure*}

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x),
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$ and $z_t = \rho_t s_t$.
Furthermore, combining eq.~\ref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the probability of fixation of phenotype $A$ when its initial frequency if $x$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citet{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection  affects only those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population is effectively masked (for better or worse) from selection at each generation.

@@

We now describe a model in which fluctuations in selection and transmission occur in space, rather then time.
Consider a population divided to two demes.
Selection (e.g. reproduction) and transmission (e.g. learning, development) occur within the demes, followed by migration of sub-adults---individuals that have already acquired their phenotype but have yet to reproduce. 
The frequency of phenotype $A$ in deme $j$ is denoted by $x_j$, and therefore after selection and transmission the frequencies are 
\begin{equation} \label{eq:migration_model_selection_transmission}
x_j^s = \rho_j \frac{w_j}{\overline{w}_j} x_j + (1-\rho_j) x_j,
\end{equation}
where $w_j$ is the fitness of phenotype $A$ in deme $j$ relative to the fitness of phenotype $B$; $\overline{w}_j=w_j x_j + (1-x_j)$ is the mean fitness in deme $j$; and $\rho_j$ is the vertical transmission rate in deme $j$.

@@

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/asym_migration.pdf}
\caption{
\textbf{Oblique transmission and asymmetric migration.}
Classification of the stable equilibrium of the system in eqs.~\ref{eq:migration_model_selection_transmission}-\ref{eq:migration_model_migration} for different ratios of the migration rates (x-axis) and vertical transmission rates (y-axis) in the two demes.
Stability was determined for 10,000 random choices of $m_1$, $m_2$, $\rho_1$, and $\rho_2$ by calculating the leading eigenvalue of the Jacobian of the system.
Blue markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=0$ was less than 1, leading to fixation of $B$.
Red markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=1$ was less than 1, leading to fixation of $A$.
Green markers denote cases in which both leading eigenvalues were larger than 1, leading to a protected polymorphism.
Here, the fitness values are $w_1=1/w_2=2$.
}
\label{fig:asym_migration}
\end{figure*}

@@

An interesting model with a similar structure was studied by \citet{Bisin2001}: it included both vertical (i.e.~``socialization in the family'') and non-vertical (i.e.~``imitation from society'') transmission.
They focused on a frequency-dependent transmission mode called \emph{cultural substitution}, in which a parent invests more in child socialization if the parental trait is rare.
Their model (eq. 12) is a continuous-time equivalent of our model (eq.~\ref{eq:recurrence}) that does not distinguish between horizontal and oblique transmission; considers constant, rather than fluctuating, selection; and includes endogenously fluctuating transmission due to changes in trait frequency.
They found that under some selection schemes, in which parents choose how many offspring to produce, either an unstable polymorphism (Proposition 4) or a stable polymorphism (Proposition 5) exists, depending on the costs associated with the transmission mode.
Furthermore, in the absence of selection trait polymorphism is neutral in our model (set $s=0$ in eq.~\ref{eq:recurrence}), but \citet{Bisin2001} found that frequency-dependent transmission leads to a globally stable polymorphism~(Proposition 1).

@@

% bibliography
\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms_sunnyvale}

>>>>

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

@@

\author[ac*]{Yoav Ram}
\author[b]{Uri Liberman}
\author[a\dagger]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{School of Mathematical Sciences, Tel Aviv University, Israel}
\affil[c]{The Interdisciplinary Center, Herzliya, Israel}
\affil[*]{{\it Current address}}
\affil[$\dagger$]{{\it Corresponding author}}

@@

Suppose that both transmission and selection fluctuate together, so that when $A$ is favored, with $w_A=1+s$ and $w_B=1$ with $s>0$, the transmission rate is $\rho_A$, and when $B$ is favored, with $w_A=1$ and $w_B=1+s$, the transmission rate is $\rho_B$.
The change in the frequency $x$ of phenotype $A$ when either $A$ or $B$ is favored is described by $F_A(x)$ or $F_B(x)$, respectively, where
\begin{equation}\begin{aligned} \label{eq:recurrence_periodic_fluc}
&F_A(x) = \rho_A \frac{1+s}{1+sx} x + (1-\rho_A)x, &\quad \text{and} \\
&F_B(x) = \rho_B \frac{1}{1+s-sx} x + (1-\rho_B)x.
\end{aligned}\end{equation}
If the environment fluctuates between favoring $A$ for $k$ generations and $B$ for $l$ generations constituting a period of $(k+l)$ generations, we describe this as $AkBl$.

@@

% Period of one
\paragraph{The case $A1B1$.}
In the case  $k=1$ we can find $x^*$ the frequency of $A$ at the protected polymorphism.
Set $F_B(F_A(x))-x = G(x) \cdot x (1-x) \cdot \gamma(x)$ with
\begin{equation} \begin{aligned} \label{eq:xstar_periodic_fluc_k=1}
G(x) &= ax^2+bx+c, \\
\gamma(x) &= -\frac{1 + sx}{s} \Big[ 1 + s + s^2 x (1-\rho)(1+x)\Big], \\
a &= s^2 \rho_A (1-\rho_A) (1 - \rho_B + \rho_B/\rho_A)) > 0, \\
b &= s(1-\rho_A)(2\rho_B - s\rho_A(1-\rho_B)), \\
c &= \rho_B - \rho_A - s\rho_A(1-\rho_B).
\end{aligned} \end{equation}
Then $x^*$ is a solution of the quadratic $G(x)=0$.
The condition \eqref{eq:poly_condition_periodic_fluc}, which guarantees that $0 < x^* < 1$, is equivalent to $c<0$, and therefore it also guarantees that $\sqrt{b^2-4ac} > b$.
So, if $0<s<1$ ($b$ is then guaranteed to be positive if LHS of eq.~\ref{eq:poly_condition_periodic_fluc} holds), then 
\begin{equation}
x^*= \frac{-b+\sqrt{b^2-4ac}}{2a}.
\end{equation}
\autoref{fig:rho1_rho2_phase_k=1} shows $x^*$ and highlights the area of the parameter space in which a protected polymorphism exists.
The figure demonstrates that the stronger the selection, the greater the fluctuations in $\rho$ that still allow a polymorphic population (area between the dashed lines).

@@

% Asymmetric period
\paragraph{General period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
& 1 < (F'_A(0))^k (F'_B(0))^l = a^k \big(b - s \Delta \rho /(1+s)\big)^l, \quad \text{and} \\
& 1 < (F'_A(1))^k (F'_B(1))^l = b^k (a + s \Delta \rho)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits a protected polymorphism; decreasing $\rho_B$  will have the opposite effect: it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Second, if $s_t$ and $\rho_t$ are not independent and $cov(\rho_t, s_t) \ne 0$, a fixation can occur.
Let $w_{A,t}, w_{B,t} \sim U(0,1)$, $s_t=(w_{A,t}-w_{B,t})/w_{B,t}$ and $\rho_t \sim Beta(1+s_t, 1)$ (a beta distribution with parameters $1+s_t$ and $1$). 
The covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 4$ as estimated by averaging over $10^8$ random values of $s_t$ and $\rho_t$); that is, vertical transmission is more likely when $A$ is favored (i.e. $s_t>0$) and oblique transmission is more  likely when $B$ is favored (i.e. $s_t<0$; \autoref{fig:beta}).
Then $E[log(1+z_t)] >0$ and $B$ almost surely doesn't fix. 
Also, $E[log(1-z_t / (1+s_t)] < 0$, so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1, 1+s_t)$ and the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -4$). In that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=\linewidth]{../figures/{beta}.png}
\caption{
\textbf{Covariance of selection and transmission.}
\textbf{(a)} Histogram of $w_{A,t}/w_{B,t}$ where $w_{A,t}$ and $w_{B,t}$ are identically and independently distributed uniform random variables $U(0,1)$.
\textbf{(b)} Histogram of $s_t = (w_{A,t}-w_{B,t})/w_{B,t}$.
\textbf{(c)} Histogram of $\rho_t \sim Beta(1+s_t, 1)$.
\textbf{(d)} The joint distribution of $\rho_t$ and $s_t$ demonstrates a positive correlation $cov(s_t, \rho_t)>0$.
}
\label{fig:beta}
\end{figure*}

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x),
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$ and $z_t = \rho_t s_t$.
Furthermore, combining eq.~\ref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the probability of fixation of phenotype $A$ when its initial frequency is $x$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citet{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection  affects only those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population is effectively masked (for better or worse) from selection at each generation.

@@

We now describe a model in which fluctuations in selection and transmission occur in space, rather then time.
Consider a population divided to two demes.
Selection (e.g. reproduction) and transmission (e.g. learning, development) occur within the demes, followed by migration of sub-adults---individuals that have already acquired their phenotype but have yet to reproduce. 
The frequency of phenotype $A$ in deme $j$ is denoted by $x_j$, and therefore after selection and transmission the frequencies $x_j^s$ are 
\begin{equation} \label{eq:migration_model_selection_transmission}
x_j^s = \rho_j \frac{w_j}{\overline{w}_j} x_j + (1-\rho_j) x_j,
\end{equation}
where $w_j$ is the fitness of phenotype $A$ in deme $j$ relative to the fitness of phenotype $B$; $\overline{w}_j=w_j x_j + (1-x_j)$ is the mean fitness in deme $j$; and $\rho_j$ is the vertical transmission rate in deme $j$.

@@

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/asym_migration.png}
\caption{
\textbf{Oblique transmission and asymmetric migration.}
Classification of the stable equilibrium of the system in eqs.~\ref{eq:migration_model_selection_transmission}-\ref{eq:migration_model_migration} for different ratios of the migration rates (x-axis) and vertical transmission rates (y-axis) in the two demes.
Stability was determined for 10,000 random choices of $m_1$, $m_2$, $\rho_1$, and $\rho_2$ by calculating the leading eigenvalue of the Jacobian of the system.
Blue markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=0$ was less than 1, leading to fixation of $B$.
Red markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=1$ was less than 1, leading to fixation of $A$.
Green markers denote cases in which both leading eigenvalues were larger than 1, leading to a protected polymorphism.
Here, the fitness values are $w_1=1/w_2=2$.
}
\label{fig:asym_migration}
\end{figure*}

@@

An interesting model with a similar structure was studied by \citet{Bisin2001}: it included both vertical (i.e.~``socialization in the family'') and non-vertical (i.e.~``imitation from society'') transmission.
They focused on a frequency-dependent transmission mode called \emph{cultural substitution}, in which a parent invests more in child socialization if the parental trait is rare.
Their model (eq. 12) is a continuous-time equivalent of our model (eq.~\ref{eq:recurrence}) that does not distinguish between horizontal and oblique transmission; considers constant, rather than fluctuating, selection; and includes endogenously fluctuating transmission due to changes in trait frequency.
They found that under some selection schemes, in which parents choose how many offspring to produce, either an unstable polymorphism (Proposition 4) or a stable polymorphism (Proposition 5) exists, depending on the costs associated with the transmission mode.
Furthermore, in the absence of selection, trait polymorphism is neutral in our model (set $s=0$ in eq.~\ref{eq:recurrence}), but \citet{Bisin2001} found that frequency-dependent transmission leads to a globally stable polymorphism~(Proposition 1).

@@

% bibliography
\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms_sunnyvale}
